# 1. Generating PyRate Input File (2 Methods)

## 1.A. CSV From Outside Source

**Loading the file** our functions will come from

```{r}
source("../pyrate_utilities.r")
```

**Download fossil occurrences** for a clade from the [Paleobiology Database](https://paleobiodb.org/classic/displayDownloadGenerator). E.g. search for the genus *Canis*

-   Check the box "Show accepted names only" in the "Select by taxonomy" section
-   Uncheck the box "Include metadata at the beginning of the output" in the "Choose output options" section. - Save it as a cvs file, e.g. using the file name *Canis_pbdb_data.csv*.

**Defining a vector** of extant (still alive) species from all the species in the data

```{r}
extant_dogs = c("Canis rufus","Canis lupus","Canis aureus","Canis latrans","Canis mesomelas","Canis anthus","Pseudalopex gymnocercus","Canis adustus","Canis familiaris")
```

**Parse raw PBDB data with extract.ages.pbdb() → generate 3 files:**

\* = name of the dataset

1.  **\*.txt** = dataset as a .txt
    1.  Then the function passes the .txt to the **extract.ages()** function, which creates the following two files
2.  **\*\_TaxonList.txt** = list of species in the dataset and their status (extinct vs. extant). This is created because the **extract.ages()** function has the argument 'save_tax_list' set to TRUE by default
3.  **\*\_PyRate.py** = PyRate formatted input, with the specified number of replicated datasets (if replicates 'argument' was used)
    1.  \*See "**example_files/Canis_example/PyRate_input_txt_visualized.png**"
    2.  In the resultant file, there is a list of arrays containing a list of numbers. The list is one PyRate input dataset, parsed form the raw data. In this example there is only one list because we didn't specify any replicated data (replicates argument below)
    3.  Each array represents one taxon (species)
    4.  Each value in the array represents one fossil occurence. Each value is a randomly sampled age (in mills of years) of the fossil occurrence for that taxon/array. Longer arrays = more occurrences of a fossil of that species, so more ages
    5.  The taxa_names variables at the end of the file = the order of the arrays
    6.  If a taxon is extant (still alive), the last value in its array is = 0 to represent the present day

```{r}
extract.ages.pbdb(file = "../example_files/Canis_example/Canis_pbdb_data.csv", extant_species = extant_dogs)
```

-   **Replicates**: The `replicates` argument specifies the number of times the age range (min and max) of each fossil occurrence should be resampled. Resampling = when the function creates that n number of datasets, in each dataset it will select a random age range for each species from within that species' original age range. Setting `replicates = 10` will generate 10 replicated datasets in the PyRate input file, each with different age ranges selected for each species. Then in the downstream PyRate analysis, you can analyze each dataset separately, and the combined results will account for age uncertainties in the fossil record.

    -   So if a fossil occurrence has a min age of 10 Ma and a max age of 12 Ma, and `replicates=10` is specified, the function will generate 10 replicated datasets where the age of that fossil occurrence is randomly sampled between 10 Ma and 12 Ma for each of the 10 replicates.

-   **Random**: The `random` argument is set to `TRUE` by default, which means that the resampling of ages will be done randomly within the specified temporal ranges.

-   **Cutoff**: The `cutoff` argument can be used to remove fossil occurrences with an age range greater than the range you specify (in millions of years). For example, setting `cutoff = 10` will remove all occurrences where the difference between the maximum and minimum age is larger than 10 million years.

### 1.A.i. TERMINAL: data summary on the \*\_PyRate.py file

Obviously, be in the PyRate folder. This command will generate summary statistics for the dataset and save those off as a .txt file with a name of your choosing. The .txt can be used in a dataframe later

`python PyRate.py 'example_files/Canis_example/Canis_pbdb_data_PyRate.py' -data_info \| tee tutorial_1_data_summary.txt`

## 1.B. CSV Created by Hand

**Prepare a fossil occurrence table** **in .txt format** with 4 columns: Taxon name, Status (extinct or extant), Minimum age, and Maximum age. An additional column can be included for a trait value (e.g., body mass)

**Loading the file** our functions will come from

```{r}
source("../pyrate_utilities.r")
```

**Parse the raw data with extract.ages() –\> generate 2 files:**

1.  **\*\_PyRate.py**
2.  **\*\_TaxonList.txt**

```{r}
extract.ages(file="../PyRate/example_files/Ursidae.txt", replicates=10)
```

'**Replicates**', '**cutoff**' arguments available in this function too.

## 1.C. Input Dataset with a "Site" Column

If your data contains info on which site a fossil came from, and some fossils were found at the same site, we will randomize the ages by the common site NOT by each fossil occurrence individually.

I.e., earlier we said extract.ages() and extract.ages.pbdb() (functions for parsing the raw data) would –\> PyRate.py input file where each fossil's age was a random sample taken from within their individual, unique temporal age range (min_age to max_age).

Now, we're saying that if we have data that specifies which site a fossil came from, and fossils share sites, we will instead assign each fossil occurrence an age that is a random sample of the SITE's unique temporal range. We'll average the min and max ages of all the fossils occurring at the same site, randomly sample within that, then assign EVERY fossil that occurred at that site the same randomly sampled site's age. So every fossil found at Site A will have the same randomly sampled age. Process:

1.  The average minimum age and maximum age of all fossils found in Site A will be calculated, and assigned as the temporal range for Site A, called 'mimima_1' and 'maxima_1' respectively
2.  The function runif() will randomly select an age from within each Site's range 'mimima_1' to 'maxima_1', round it to 6 decimal places, and store it as a new column (new entry in each array in \*\_PyRate.py) '**new**\_**age**'
3.  The resultant \*\_PyRate.py file will look the same as "**example_files/Canis_example/PyRate_input_txt_visualized.png**" except that each array (taxon/species) will only have a maximum of two values:
    1.  One value corresponding to its site's randomly sampled age.
    2.  A 0 value at the end of the array, if it's an extant species (discussed previously)

## 1.D. Check \*\_TaxonList.txt for Species Name Typos

**TERMINAL** command:

`python PyRate.py -check_names *_TaxonList.txt`

This will return a table ranking possible typos. You should check these yourself to see if they are errors. If they are, go back to the dataset, fix it, and re-run the needed steps above.

-   Ranks 0-1 = possible misspellings

-   Ranks 2-3 = possibly different names

# 2. Estimation of Speciation & Extinction Rates Through Time

## 2.A. Defining the Preservation Model Type in TERMINAL

### 2.A.i.  (NHPP) Non-Homogenous Poisson Process of Preservation

NHPP models the preservation rate as a function of time, which = the preservation process can vary (has temporal heterogeneity) over time

```{python}
# Function in PyRate.py to calcculate the likelihood of the NHPP AND the TPP model
NHPP_lik()
```

### 2.A.ii. (HPP) Homogenous Poisson Process

HPP is the simplest model. It assumes preservation rate is constant: a single parameter. In **Terminal**, feed your PyRate input file into PyRate, and run the HPP model.

`python PyRate.py *_PyRate.py -mHPP`

```{python}
# The HOMPP_lik function in PyRate.py calculates the likelihood of the HPP model.
HOMPP_lik()
```

### 2.A.iii. (TPP) Time-Variable Poisson Process

In the TPP model, the preservation rates are allowed to vary across different epochs or time intervals that the USER DEFINES. Each epoch has its own preservation rate, and each individual preservation rate has its own prior probability distribution. The model will output different preservation rates for each interval.

You need to create a file (below called epochs_q.txt) specifying the time intervals (epochs).

In **Terminal**, feed your PyRate input file + a file with times delimiting the epochsinto PyRate, and run the TPP model.

`python PyRate.py *_PyRate.py -qShift epochs_q.txt`

So we have a preservation rates vector (the list of preservation rate occurring for each epoch we specified). The prior probability on that vector = a probability distribution PER preservation rate in each epoch = the probability distribution of that preservation rate ocurring before considering the observed fossil occurrence data

We need to set this prior probability distribution for each preservation rate. We model them each as their own gamma distribution w/ two parameters: shape and rate. The default is shape = 1.5 and rate = 1.5. This default prior assumes preservation rates are more likely to be concentrated around moderate values, with lower probability of extremely high or low rates.

-   Shape \> 1 = unimodal distribution

-   Shape \< 1 = exponentially decreasing distribution

-   Rate = scaling the gamma distribution = variability in preservation rates = changing the mean and variance of the preservation rates

    -   Larger rate = distribution concentrated around smaller values

    -   Smaller rate = more spread out distribution

    -   1.5 is a moderate variability in preservation rates

Being able to change the gamma distribution means we have flexibility in modelling the shape/scale of prior probability

Adding `-pP shape rate` to the end of that command is optional. The default prior on the vector of preservation rates is a single gamma distribution with shape parameter = 1.5 and rate parameter = 1.5. These values can be changed using the `-pP` command, e.g., `-pP 2 0.1`.
The rate parameter of the prior can be estimated from the data by setting it to 0.

During the MCMC sampling process, PyRate explores different values for the preservation rates in each epoch, guided by the prior probability distributions. The posterior distribution of the preservation rates is then obtained by combining the prior probabilities with the likelihood of the observed fossil occurrence data.

By using a probability distribution as the prior for each preservation rate, PyRate allows for uncertainty and variability in the estimated rates across different epochs. The gamma distribution provides a flexible framework to express prior beliefs about the range and shape of the preservation rate distributions, while still allowing the data to inform the posterior estimates.

The model will output different preservation rates for each interval. These are known as the posterior probability (P(A\|B) = our model's updated belief about the preservation rates after we considered both

-   P(A) the prior probabilities (gamma distributions of the preservation rate in each epoch)

-   and P(B\|A) the likelihood of observing the data (fossil occurrences)

```{python}
# Variable to indicate whether TPP model is being used. == 1 means yes
TPP_model
```

### 2.A.iv. Gamma Model: Accounting for Preservation Rate Heterogeneity

Can be coupled with NHPP, HPP, TPP

## 2.B. Analysis of Model
