# 1. Generating PyRate Input File (2 Methods)

## 1.A. CSV From Outside Source

Loading the file our functions will come from

```{r}
source("../pyrate_utilities.r")
```

Defining a vector of extant dog species

```{r}
extant_dogs = c("Canis rufus","Canis lupus","Canis aureus","Canis latrans","Canis mesomelas","Canis anthus","Pseudalopex gymnocercus","Canis adustus","Canis familiaris")
```

Parse raw PBDB data --\> generate 2 files:

1.   \*\_SpeciesList.txt = list of all species names
2.  \*\_PyRate.py = PyRate formatted input

```{r}
extract.ages.pbdb(file = "../example_files/Canis_example/Canis_pbdb_data.csv", extant_species = extant_dogs)
```

### 1.A.i. TERMINAL: data summary on the \*\_PyRate.py file

Obviously, be in the PyRate folder. This command will generate summary statistics for the dataset and save those off as a .txt file with a name of your choosing. The .txt can be used in a dataframe later

python PyRate.py 'example_files/Canis_example/Canis_pbdb_data_PyRate.py' -data_info \| tee tutorial_1_data_summary.txt

## 1.B. CSV Created by Hand

Prepare a fossil occurrence table with 4 columns: Taxon name, Status (extinct or extant), Minimum age, and Maximum age. An additional column can be included for a trait value (e.g., body mass)

Loading the file our functions will come from

```{r}
source("../pyrate_utilities.r")
```

Parse the raw data and generate PyRate input file

extract.ages(file="../PyRate/example_files/Ursidae.txt", replicates=10)

```{}
```
